---
title: "McKenzie Metadata Management"
output:
  html_document:
    df_print: paged
    code_folding: hide
    toc: true
    toc_float: true
    toc_collapsed: false
---

```{r, message=FALSE, warning=FALSE}
require(tidyverse)
require(knitr)
require(magrittr)
require(lubridate)
require(knitr)
```

# Rationale

Releases of Chinook salmon on the McKenzie River are from two sources: hatchery outplants and the Cougar Trap. Hatchery outplants are fish that volitionally enter hatcheries and are trucked to release sites above the dam. Cougar trap fish include both NOR and HOR individuals that enter the trap at the base of the Cougar Dam and are either released downstream, into the tailrace or above the dam. Cougar trap metadata is already organized in the notebook in this directory ("cougar_trap_metadata_mgmt.Rmd") and saved to file ("cougar_consolidated_metadata.xslx.). 

There are also two additional groups of samples: precocial males (of unknown provenance), and individuals sampled in spawning ground surveys. 

In this notebook we import and clean up metadata from the hatchery outplants, spawning ground surveys and precocial males, and then combine with metadata from the Cougar Trap into into a single, consolidated dataframe and file to be used for future analyses.

We also compare the metadata to summary outputs from previous reports to ensure it's accuracy and create a log of all unresolved questions and issues.

# Import Datasets

First let's import data. 

```{r, cache = TRUE, message=FALSE, warning=FALSE}
# previous data

# there are several hundred files called "cougar_adult_master", this is the latest and most accurate version according to Nick (email correspondence Nov 9 2021)
sard_2016 <- readxl::read_xlsx("../../prev_metadata/10APR2016_Cougar_adult_master.xlsx" , sheet = 1, guess_max = 6593) 

# there are also some data from Vickie zeller here:
# South Fork McKenzie_Cougar Dam_Master_2007-2017_Data.xlsx
# we hold off on these for now since they are split across >20 tabs and I think all the most useful information has already been extracted for the Cougar dataset

# Sandra Progeny - Sandra Bohns progeny query Sept 2020
bohn_2020 <- readxl::read_xlsx("../../prev_metadata/Ots McKenzie Basin progeny export 2020-09-23.xlsx" , sheet = 1, guess_max = 1000) 

# dayan progeny
progeny_2021 <- readxl::read_xlsx("../../dayan_metadata/processed_metadata/Dayan_Oct2021_progeny_export.xlsx" , sheet = 1, guess_max = 6593) 

# cougar consolidated
cougar <- readxl::read_xlsx("cougar_consolidated_metadata.xlsx", sheet = 1, guess_max = 2500)

```

# Release Sites

rkm == 13 ~ "Slide Creek"
rkm == 17.2 ~ "Hard Rock"
rkm == 18.5 ~ "Bridge 1980"
rkm == 29.7 ~ "Homestead" / Bridge 430
rkm == 36.4 ~ "Frissel Bridge"

# Hatchery Outplants

Hatchery outplants in this document refers to individuals that were hauled from the hatchery (Leaburg of McKenzie) and outplanted above Cougar Dam onthe South Fork McKenzie. They do not include HOR individuals transported from the Cougar trap, and they do not include individuals from the hatchery outplanted in other locations in the McKenzie basin such as above Trailbridge dam

## 2007 - 2013: Sard

From 2007 to 2013, we should go with Nick's data for hatchery outplants.

```{r}
# let's revise sard's data to match the final format
sard_revised_outplants <- sard_2016 %>%
  rename(OldLabCode = Sample.Name) %>% #rename to match progeny field for merging
  filter(out.strat == "hatch")# we're only interested in the metadata from the hatchery outplants for now
  
# there are some fields in the progeny outplut not in the sard  

# let's collect the SFGL ids from Sandras data

#sard_revised_outplants$OldLabCode[!(sard_revised_outplants$OldLabCode %in% bohn_2020$OldLabCode)]

# 2 fish don't match, let's fix this prior to merging
ids_a <- bohn_2020 %>%
  select(sample_id = `Individual name`, OldLabCode, date = DateSampled) %>%
  mutate(OldLabCode = case_when(sample_id == "OtsAC13SFMK_4932a" ~ "MRA13_4932.a",
                               sample_id == "OtsAC13SFMK_4932b" ~ "MRA13_4932.b",
                               TRUE ~ OldLabCode
                                 )) # Progeny data entry error (MRA13_4932.a instead of MRA13_4932a )

sard_revised_outplants %<>%
  left_join(ids_a)

# now let's clean up and make consistent with the cougar dataset
sard_revised_outplants %<>%
  select(sample_id, origin = type, geno_sex = sex.gt, pheno_sex = sex.pheno, rkm, date, year, length) %>%
  mutate(origin = case_when(origin == "H" ~ "HOR",
                            TRUE ~ origin),
         geno_sex = case_when(geno_sex == "0" ~ NA_character_,
                              geno_sex == "RR" ~ NA_character_,
                              TRUE ~ geno_sex),
         pheno_sex = case_when(pheno_sex == "UK" ~ NA_character_,
                               TRUE ~ pheno_sex),
         date = ymd(date),
         release = case_when(rkm == 13 ~ "Slide Creek",
                             rkm == 17.2 ~ "Hard Rock",
                             rkm == 18.5 ~ "Bridge 1980",
                             rkm == 29.7 ~ "Homestead",
                             rkm == 36.4 ~ "Frissel Bridge")) # these are inferred from the Zeller metadata file
```

## 2014 - 2015: 

2014 - 2015 is where things get messy.
Can we use the Bohn progeny query for these? Or do we need to look elsewhere?

Let's go year by year.

__2014__
Tissue audit (see file sample_summary.docx in this repository) suggests there are 561 hatchery outplants from 2014, but only 491 are possible parents, as 70 were actually release above the Trailbridge dam, not the Cougar Dam.
```{r}
bohn_2020 %>%
  filter(str_starts(`Individual name`, "OtsAC14MCKR")) %>% 
  count(ReleaseLocation)
```

The Bohn dataset is missing the release location, so we need to look elsewhere.

```{r}
progeny_2021 %>%
  filter(str_starts(`Individual name`, "OtsAC14MCKR"), Marks == "Clipped") %>% 
  count(RiverLocation) # whoever entered data used the RiverLocation field, not ReleaseLocation

# YES this works, use this one
```

The 2016 technical report states there are 689 total hatchery origin outplants, including Cougar Trap samples. Does this add up?

```{r}
cougar %>%
  filter(year == 2014, origin == "HOR") %>%
  nrow()
```

No. I've dedicated many hours trying to solve this discrepancy at many points in the past year. 


__2015__  
Tissue audit (see file sample_summary.docx in this repository) suggests there are 600 hatchery outplants in 2015 (600 hatch outplants + 19 Cougar Trap HOR outplants). This agrees with everything else we have (Vickie raw files, progeny exports, 2016 tech report) except the value Sandra wrote in the proposal (669). Never explained where the extra 50 individuals come from. Especially confusing is that in all other cases Sandra lumped trailbridge releases with cougar releases, so why arent the additional 100 trailbridge outplants included? Like the 2014 issue above spent many hours trying to resolve this one.

```{r}

progeny_2021%>%
  filter(str_starts(`Individual name`, "OtsAC15MCKR")) %>% 
  count(RiverLocation)

# yes filtering the 2021 progeny export for hard rock and 430 bridge will get us all we need
```

## 2016 and 2017
2016 was batch sampled so there is no individual level information in progeny. Instead there are 5 batch jars of fin clips from outplants above Cougar, separated by date of release and sometimes by sex. Manually created a spreadsheet of the available information for individual fin clips with applicable metadata and used that for 2016 and 2017.

```{r}
batch_hatch <- readxl::read_xlsx("Mckenzie_Batch_Sampling.xlsx", sheet = 2)
```

## Combine and QC

Now let's put these together and reformat to match the Cougar dataset

__2007 - 2013__
Still a few small schanges to make

```{r}
sard_revised_outplants %<>%
  mutate(length = as.numeric(length))
```


__2014__
2014 data is still incomplete, a little more work below
```{r}
hatch_out_2014 <- progeny_2021 %>%
  filter(str_starts(`Individual name`, "OtsAC14MCKR"), Marks == "Clipped", RiverLocation %in% c("Hardrock Campground", "430 Bridge", "Frissel Crossing"))  %>%
  select(sample_id = `Sample Name`, pheno_sex = Gender, length = LengthFork1, release = RiverLocation) 

# agh my progeny export grabbed SampleDate and ReleaseDate, but someone uploaded the date to the field DateSampled! progeny truly a nightmare...luckily the Sandra export got DateSampled (Even though it missed the release location field...), together we have everything

hatch_out_2014 %<>%
  left_join(select(bohn_2020, "Individual name", DateSampled), by = c("sample_id" = "Individual name")) %>%
  mutate(origin = "HOR") %>% 
  mutate(release = case_when(release == "Hardrock Campground" ~ "Hard Rock",
                             release == "Frissel Crossing" ~ "Frissel Bridge",
                             release == "430 Bridge" ~ "Homestead",
                             TRUE ~ release),
         date = lubridate::as_date(DateSampled),
         year = lubridate::year(date)) %>%
  select(-DateSampled)

```

__2015__
2015 missing the same info as 2014, but 2015 requires a slightly different appraoch from 2014

```{r}
hatch_out_2015 <- progeny_2021 %>%
  filter(str_starts(`Individual name`, "OtsAC15MCKR"), Marks == "Clipped", RiverLocation %in% c("Hard Rock", "430 Bridge"))  %>%
  select(sample_id = `Sample Name`, pheno_sex = Gender, length = LengthFork1, release = RiverLocation) 

# agh my progeny export grabbed SampleDate and ReleaseDate, but someone uploaded the date to the field DateSampled! progeny truly a nightmare...luckily the Sandra export got DateSampled (Even though it missed the release location field...), together we have everything

hatch_out_2015 %<>%
  left_join(select(bohn_2020, "Individual name", DateSampled), by = c("sample_id" = "Individual name")) %>%
  mutate(origin = "HOR") %>% 
  mutate(release = case_when(release == "430 Bridge" ~ "Homestead",
                             TRUE ~ release),
         date = lubridate::as_date(DateSampled),
         year = lubridate::year(date)) %>%
  select(-DateSampled)

```

__2016 and 2017__

```{r}
batch_hatch %<>%
  mutate(date = lubridate::as_date(date),
         origin = "HOR",
         year = lubridate::year(date))
```

__Combine__


```{r}
hatch_outplants <- bind_rows(sard_revised_outplants, hatch_out_2014, hatch_out_2015, batch_hatch)
```

Yay, almost complete.

We still need to add the genotypic sex to individuals freshly genotyped.

# Spawning Ground Surveys

Next let's get the SGS metadata. SGS are mostly used to add potential NOR offspring to the dataset that aren't sampled at the trap, but in a few years SGS were conducted above the dam and need to be included as potential parents.

Also note that there are many more SGS samples than we include as potential offspring. We only include SGS from the confluence of the mainstem to the base of Cougar Dam

SGS are conducted from 2011-2019

## 2011 - 2015

For 2011 to 2015, we can rely on Nicks data. Here from a different source.

```{r}
sard_sgs_raw <- read_tsv("../../prev_metadata/04JAN2016_below_dam_Cougar_adult_master.txt")
reaches <- read_tsv("../raw_metadata/spawning_ground_surveys/survey_reaches.txt")

sard_sgs <- sard_sgs_raw %>%
  left_join(reaches, by = c("loc" = "reach")) %>% 
  filter(section %in% c("SFMK_above", "SFMK_below")) %>%
  select(hmsc_id = "Sample Name", geno_sex =sex.gt, pheno_sex = sex.pheno, sgs_reach = loc, date = jul_date, year, sgs_above_below = section) %>%
  mutate(origin = "NOR") %>% 
  left_join(select(progeny_2021, sample_id = "Individual name", OldLabCode ), by = c("hmsc_id" = "OldLabCode"))

#clean up

sard_sgs %<>%
  mutate(sample_id = case_when(is.na(sample_id) ~ hmsc_id,
                               TRUE ~ sample_id),
         geno_sex = case_when(geno_sex== "UK" ~ NA_character_,
                              TRUE ~ geno_sex),
         pheno_sex = case_when(pheno_sex == "Female" ~ "F",
                               pheno_sex == "Male" ~ "M",
                               TRUE ~ pheno_sex),
         date = lubridate::mdy(date))
  
```

2011-2013 do not have SFGL ids and are not in progeny, we will use the original hmsc id as the sample name for these, 2014-2015 will use the SFGL ids

## 2016 - 2019

The progeny fields for 2016-2019 are a total mess. Let's start from the raw metadata, because this is one of the few places where we have it.

__2016__
```{r}
sgs_2016 <- read_tsv("../raw_metadata/spawning_ground_surveys/McKenzie_2016_CarcassSurvey.txt")

reaches <- read_tsv("../raw_metadata/spawning_ground_surveys/survey_reaches.txt")
sgs_2016 %<>%
  left_join(reaches, by = c("Section" = "reach")) %>%
  filter(section != "mainstem") %>%
  select(sample_id = "SFGL ID", sgs_reach = Section, section, date = Date, origin = Adipose, pheno_sex = Sex)

sgs_2016 %<>%
  mutate(date = mdy(date),
         year = 2016)


```

__2017__
```{r}
sgs_2017 <- read_tsv("../raw_metadata/spawning_ground_surveys/Upper Willamette SGS samples_2017.txt")

sgs_2017 %<>%
  filter(`Stream Name` == "South Fork McKenzie") %>%  # all are from below dam this year
  select(sample_id = `Individual Name`, date = DateSampled, origin = Marked, pheno_sex = Gender, sgs_reach = IndividualSampleLocation) %>%
  mutate(section = "SFMK_below")

sgs_2017 %<>%
  mutate(date = mdy(date),
         year = 2017)
```

__2018__

```{r}
sgs_2018 <- readxl::read_xlsx("../raw_metadata/spawning_ground_surveys/SGS_2018_combined.xlsx", sheet =1 )


sgs_2018 %<>%
  filter(Stream == "South Fork McKenzie") %>%
  select(sample_id = `Individual Name`, date = DateSampled, origin = Clipped, pheno_sex = Gender, sgs_reach = IndividualSampleLocation) %>%
  mutate(section = "SFMK_below")
  

sgs_2018 %<>%
  mutate(date = ymd(date),
         year = 2018)
```

__2019__
```{r}
sgs_2019 <- read_tsv("../raw_metadata/spawning_ground_surveys/DRAFT NAI-EAS 2019 Carcass Survey Data 04_NOV_19.xlsx - 2019 Survey Data.tsv")

sgs_2019 %>%
  filter(River == "McKenzie") %>%
  count(Reach)# check mckenzie for sfmk, NO SFMK under McKenzie!

sgs_2019 %<>%
  filter(River == "South Fork McKenzie") %>%
    select(sample_id = SFGL_ID, date = `Survey Date`, origin = `Clipped (Y/N)`, pheno_sex = `Sex (M/F)`, sgs_reach = Reach) %>%
  mutate(section = "SFMK_below") 

sgs_2019 %<>%
  mutate(date = mdy(date),
         year = 2019)
```

## Combine
```{r}
sard_sgs %<>%
  rename(section = sgs_above_below)
  
sgs_meta <- bind_rows(sard_sgs, sgs_2016, sgs_2017, sgs_2018, sgs_2019)
```


# Precocial Males

also add the precocial males

# Combine
```{r}
cougar %<>%
  mutate(date = ymd_hms(date),
         date = ymd(date),
         length = as.numeric(length))
metadata <- bind_rows(cougar, hatch_outplants, sgs_meta)
```



# Final Check

```{r}
dayan_genotyped <- readxl::read_xlsx("../../../lab_work/genotyping/to_genotype.xlsx", sheet = 2)

prog_micro <- read_tsv("../../../genotypes/genotype_data/old_genotypes/Progeny_microsats.txt")

genotyped <- c(dayan_genotyped$Sample, prog_micro$`Progeny name`, sard_sgs_raw$`Sample Name`)

missing_from_genos <- metadata %>%
  filter(!(sample_id %in% genotyped))
```

Most missing are from a change in naming convention: there are 120 2018 cougar trap samples named OtsAC18MCKR_ that should be named OtsAC18SFMK_. These are already genotyped.

All others make sense (e.g. 2019 HORs outplanted from Cougar trap - 2019 is not a parent year, so these aren't included ) except the following:

OtsAC10SFMK_1118 -   
4 2016 carcass samples: OtsCC16MCKR_1100, OtsCC16MCKR_1161, OtsCC16MCKR_1162, OtsCC16MCKR_1243 -  These got skipped when I didn't understand how to parse the different "bridge to mouth" carcass samples. They should be included. At this point (when this notbeook is being written) first round genotyping is complete and re-runs are plated, but not extracted, amplified or genotyped.

Fortunately there are several spots on the re-run plates where all the tissue was used in the first round genotyping, so no re-runs are possible. I manually edited the final plating files to add these samples to the re-runs.

