---
title: "McKenzie Genotype Comparison"
output:
  html_document:
    df_print: paged
    code_folding: hide
    toc: true
    toc_float: true
    toc_collapsed: false
---

```{r, message=FALSE, warning = FALSE}
require(tidyverse)
require(magrittr)
require(knitr)
```


# Rationale

Here we compare microsatellite genotype scoring between David Dayan and Dave Jacobsen. Genotypes are scored from the same ABI output files, using the same genemapper alleles/bins. The only difference is who is doing the scoring.

# Data Prep

Let's get the data into R and into a format that allows for comparison.

## Import

First, let's get DPJ's genotypes in.

```{r, message=FALSE, warning = FALSE}
dj_1 <- readxl::read_xlsx("DavidD_Comp_Pt5-8_Panels1-3.xlsx", sheet =1)
dj_2 <- readxl::read_xlsx("DavidD_Comp_Pt5-8_Panels1-3.xlsx", sheet =2)
dj_3 <- readxl::read_xlsx("DavidD_Comp_Pt5-8_Panels1-3.xlsx", sheet =3)
```

Now DID's

```{r, message=FALSE, warning=FALSE}
dd <- read_tsv("MCKR_Plate_5-8 Genotypes Table.txt", guess_max = 5000)
```

## Clean Up

Now le't start making these tables tidy.

Let's have one row for each genotype in one big dataframe.

Start by cleaning up DID

```{r}
dd %<>%
  select(sample = `Sample Name`, panel = Panel, marker = Marker, a1_did = `Allele 1`, a2_did = `Allele 2`)

# also made te mistake of leaving an uncalled allele in there a lot, e.g. meant to call as a homozygote, but there's a ? allele still included.
# let's convert these two homozygotes

dd %<>%
  mutate(a1_did = case_when(a1_did == "?" & a2_did != "?" ~ a2_did,
                            TRUE ~ a1_did)) %>%
  mutate(a2_did = case_when(a2_did == "?" & a1_did != "?" ~ a1_did,
                            TRUE ~ a2_did))

```

Now DPJ
```{r}
# first get rid of unused columns, then pivot longer
dj_1 %<>%
  select(-c(`Sample File`, Lane, Well)) 

#theres multiple things we have to resolve here
# first colnames can't be redundant, and we want each genoype to appear on its own row, not each allele. so here's a strategy to successfully make this dataset long:

# 1 combine columns that start the same
# pivot longer 
# split GT column

dj_1 %<>%
  unite("201", "201...6", "201...7") %>%
  unite("209", "209...8", "209...9") %>%
  unite("409", "409...10", "409...11") %>%
  unite("515", "515...12", "515...13") %>%
  pivot_longer(cols = -c(`Sample Name`, Panel), names_to = "marker", values_to = "gt") %>%
  separate(gt, into = c("a1_dj", "a2_dj"))

dj_2 %<>%
  select(-c(`Sample File`, Lane, Well)) %>% 
  unite("211", "211...6", "211...7") %>%
  unite("253b", "253b...8", "253b...9") %>%
  unite("208", "208...10", "208...11") %>%
  unite("249", "249...12", "249...13") %>%
  pivot_longer(cols = -c(`Sample Name`, Panel), names_to = "marker", values_to = "gt") %>%
  separate(gt, into = c("a1_dj", "a2_dj"))

dj_3 %<>%
  select(-c(`Sample File`, Lane, Well)) %>% 
  unite("212", "212...6", "212...7") %>%
  unite("215", "215...8", "215...9") %>%
  unite("311", "311...10", "311...11") %>%
  pivot_longer(cols = -c(`Sample Name`, Panel), names_to = "marker", values_to = "gt") %>%
  separate(gt, into = c("a1_dj", "a2_dj"))

dj <- bind_rows(dj_1, dj_2, dj_3)


```

## Combine

First why do the number of genotypes vary between DD and DJ

```{r}
miss_inds <- dd$sample[(!(dd$sample %in% dj_3$`Sample Name`))] %>% unique()

#None of DPJ's datasets include controls, and dataset 3 skips 5 individuals. Let's remove these from DD

dd %<>%
  filter(!(sample %in% miss_inds))
```

None of DPJ's datasets include controls, and dataset 3 skips 5 individuals from panel 3. Let's only compare scored individuals and remove these.

Now let's combine.

```{r}
# fix names 

dj %<>%
  rename(sample = `Sample Name`, panel = Panel)

df <- left_join(dd, select(dj, -panel), by = c("sample" = "sample", "marker" = "marker"))
```

# Compare

__diploid comparisons__   
Comparisons can be a little complicated. For example, if DD scored an individual as alleles a and b, but DJ scored as a and c, should we call this a disagreement? 
Similarly we can just compare DD allele 1 to DJ allele 1 because they are presented in order by genemapper. For example, consider a marker with alleles 1, 2, 3 and 4. If DD scored the individual as 1/3 and DJ scored as 3/4, scoring allele by allele would miss that there is partial agreement. 

Instead we need to sum the number of alleles in common for each genotype. If both alleles agree this is scored as 100% agreement. If only one, then 50%.

__missing__  
If one scorer calls an individual as missing and the other calls a genotype we count this as a disagreement, but not this will not have as strong effect on parentage assignment. 

```{r}
df %<>%
  filter(marker != "OtY3") %>%
  replace_na(list(a1_did = 0, a2_did = 0)) %>%
  rowwise() %>%
  mutate(agree_count = sum(c(a1_did, a2_did) %in% c(a1_dj, a2_dj))) %>%
  ungroup()
```

## Results

__Overall Comparison__
```{r}
sum(df$agree_count)/(2*nrow(df))
```

Overall agreement rate is 96%

__Per Marker__
```{r}
kable(df %>%
  group_by(marker) %>%
  summarise(aggreement = sum(agree_count)/(2*n())), digits = 3)

```

Three markers seem to cause most of the problems: 311 (~15% disagreement), 215 (~10% disagreement), and 212 (6% disagreement). All other markers at least 98% agreement.

__Uncalled__  
Is this due to true disagreements or one of either DD/DJ being more conservative.

```{r}
df %<>%
  rowwise() %>%
  mutate(did_uncalled = sum("0" %in% c(a1_did, a2_did))) %>%
  mutate(dj_uncalled = sum("0" %in% c(a1_dj, a2_dj))) %>%
  mutate(uncall_agree = did_uncalled == dj_uncalled)

```

Who is more conservative?
```{r}
kable(df %>%
        ungroup() %>%
        group_by(marker) %>%
        summarise(uncalled_rate_did = mean(did_uncalled),uncalled_rate_dj = mean(dj_uncalled) ), digits = 3)
```

Let's make this easier to understand: what proportion of disagreements is due to one of calling a genotype when the other doesn't?

```{r}
kable(df %>%
  group_by(marker) %>%
  summarise(disagreement = 1-sum(agree_count)/(2*n()), dis_due_to_uncalled = 1-sum(uncall_agree)/n()) %>%
  mutate(prop_dis_due_to_uncalled = dis_due_to_uncalled/disagreement), digits = 3)
```


Dave is more conservative, but only at the troubling markers (215, 212 and 311)

# Conclusion

```{r, include = FALSE}
df %>%
  ungroup() %>%
  filter(!(marker %in% c("212", "215", "311"))) %>%
  summarise(mean_agree = mean(agree_count)/2)


df %>%
  ungroup() %>%
  filter(!(marker %in% c("212", "215", "311"))) %>%
  summarise(mean_agreebc_unc = sum(uncall_agree))

```


For 8 of the 11 markers in the panel, genotypes scored by DID and DPJ demonstrate few differences (99.0% agreement), with almost all disagreements stemming from one scorer calling genotypes, when the other did not (e.g. score as missing genotype). There was not a systematic bias towards either DID or DPJ calling individuals or not calling individuals at these 8 markers.

For the remaining three markers (311, 215 and 212), disagreements were frequent:  
311 (~15% disagreement)  
215 (~10% disagreement)  
212 (~6% disagreement)  

Nearly all of the disagreement at these markers was due to DPJ not calling a genotype when DID did score a genotype. 

