---
title: "Reruns"
output:
  html_document:
    df_print: paged
    code_folding: hide
    toc: true
    toc_float: true
    toc_collapsed: false
---

```{r, message=FALSE, warning = FALSE}
require(tidyverse)
require(magrittr)
require(knitr)
```


Get previous data
```{r}
mckr <- readxl::read_xlsx("~/FRA/McKenzie_pedigree/mckenzie_reintroduction/previous_run_info/10APR2016_Cougar_adult_master.xlsx", sheet = 1)

```

Is the dataset above the filtered or raw?
```{r}
mckr %>%
  rowwise() %>%
  mutate(zero_count = sum((cur_data() == "0"), na.rm = TRUE)) %>%
  ungroup() %>%
  summarise(mean_zero = mean(zero_count), max = max(zero_count))

mckr %>%
  rowwise() %>%
  mutate(zero_count = sum((cur_data() == "0"), na.rm = TRUE)) %>%
  ungroup() %>%
  summarise(number_w_less_than_7_gts = sum(zero_count > 5))
```


# How many in plates 5-8 need reruns?

```{r}
dd %>%
  filter(marker != "OtY3") %>%
  ungroup() %>%
  group_by(sample) %>%
  summarise(na_count_greater_than_7 = sum(is.na(a1_did))) %>%
  filter(na_count_greater_than_7 >= 4)
  
```

About 1.1% of samples need re-runs for plates 5-8. These represent some of the oldest tissue samples and ones I did with the least experience working with the microsat protocol in the lab, so it's hard to imagine it gets much worse for than this (except for carcass samples). 

So the total predictd number of individuals that need to be re-run is ~29. We can assume that more than this will actually need to be re-run because carcass samples are worse, and the sex marker isn't included in this count. 

Nick's final dataset had 99% coverage at 10 or more loci. How does our dataset compare using this metric?


```{r}
dd %>%
  ungroup() %>%
  group_by(sample) %>%
  summarise(na_count_greater_than_7 = sum(is.na(a1_did))) %>%
  filter(na_count_greater_than_7 >= 2)


  
```

Only 90% of our samples are successfully genotyped at at least 10 loci. 

This is odd because DPJ told me he'd only do re-runs to reach the cutoff required for the analysis, not to attempt to complete the dataset. My guess is that this means he ran individuals that failed to reach >=7 markers, and this got the dataset up to that 99% at >= 10 marker level of completion.

What about going by DPJ's more conservative scoring?

```{r}
dj %>%
  filter(a1_dj == "0") %>%
  count(sample) %>%
  filter(n>3) %>%
  nrow(.)/length(unique(dj$sample))
```

If we go by Dave's scoring, we need to do re-runs at ~2% of samples, or about 57 individuals.

__Carcass Samples__

Plates 5-8 don't have any carcass samples.



